---
title: "Data Mining HW 3"
author: "Asha Christensen, Pranjal Maheshka, Marco Navarro"
date: "2023-03-14"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(ggplot2)
library(modelr)
library(mosaic)
library(psych)
library(rsample)  
library(caret)
library(parallel)
library(foreach)
library(gamlr)
library(verification)
library(groupdata2) 
library(rpart)
library(randomForest)
```
## What Causes What?

1. Cities with high levels of crime are incentivized to hire more cops: this means that even if cops deter crime, regressing "crime" on "police" may give an opposite result, since crime and police are positively correlated. Therefore, we need to use a different method to find the true effect of police on crime.

2. They found times in which there was a high police presence on the street which was not due to street crime: the terror alert level being orange or red. When this is the case, there is an extra police presence that is due to terrorism but not due to the type of crime they wanted to study. In this way, they could look at the effect of increased police presence on these days on street crime rates. 

The table shows that high alert terrorism days decreased the number of daily crimes by 7.316 crimes, and this effect was still large and significant after controlling for metro mid-day ridership.

3. There are 2 potential effects of a high terrorism alert (i) there is an increased police presence on the street, (ii) the terrorism threat reduces tourism and therefore potential victims on the street. Both of these could reduce the daily crime rate, so in order to isolate the first effect, we need to control for the third. Therefore, they use Metro mid-day ridership as an indicator of the number of people out-and-about. Controlling for this should isolate the effect of increased police presence on daily crime rates.


4. When we separate out crimes that occur in the national mall versus the rest of the city, we find that the reduction in crime due to high alert days primarily takes place in the national mall-- this may be due to the fact that the majority of increased police presence during a terrorism alert is concentrated on the national mall since these are the most likely terrorism threats. The decrease in crimes for the rest of the city is not statistically significant, whereas the reduction in crimes in the national mall is 2.62 crimes.


## Predictive model building: green certification

```{r, echo=FALSE}
greendf = read.csv("C:\\Users\\ashac\\OneDrive\\Documents\\GitHub\\eco395m-sp23-DataMining\\Homework-3\\greenbuildings.csv", as.is = T, fill = T, header = T)
#change this during upload? don't know how to run it tbh

greendf = greendf %>%
  mutate(revsqftyr = Rent*leasing_rate)

green_split = initial_split(greendf, prop = 0.8)
green_train = training(green_split)
green_test = testing(green_split)

```

Let's start with some step-wise models, using some medium linear models. Here are our initial out-of-sample RMSEs for both linear models:

``` {r, echo=FALSE}

lm1 = lm(revsqftyr ~ green_rating + size + age + renovated + amenities + City_Market_Rent, data=green_train)

lm2 = lm(revsqftyr ~ green_rating + cd_total_07 + hd_total07 + Gas_Costs + Electricity_Costs, data=green_train)

rmse(lm1, green_test)
rmse(lm2, green_test)

```
In these step-wise regressions, we allow the use of all variables except rent, leasing rate, property ID, LEED and Energystar, allowing for second-level interactions.

``` {r, echo=FALSE, include=FALSE}

step1 = step(lm1, scope=~(.-Rent - leasing_rate - CS_PropertyID -LEED - Energystar)^2)

step2 = step(lm2, scope=~(.-Rent - leasing_rate - CS_PropertyID -LEED - Energystar)^2)

```

```{r, echo=FALSE}
rmse(step1, green_test)
rmse(step2, green_test)

```
There's a slight improvement on out-of-sample RMSE, but not a large decrease. We may be able to use a random forest model to more accurately predict revenue per square foot per year.


```{r, echo=FALSE, cache=TRUE}

green_train_noNA = na.omit(green_train)

revenue.forest = randomForest(revsqftyr ~ (.-Rent - leasing_rate - CS_PropertyID -LEED - Energystar), data=green_train_noNA, importance = TRUE)

```

```{r, echo=FALSE}

green_test_noNA = na.omit(green_test)

modelr::rmse(revenue.forest, green_test_noNA)

```

Our out-of-sample RMSE is much lower than the linear step models produced, so this is our best model to predict revenue per square foot per year. We can now look at the partial effect of green rating.

```{r, echo=FALSE}
varImpPlot(revenue.forest)

```

According to the variable importance plot, this indicates that green rating is not one of the more important variables in predicting revenue per square foot per year.

```{r, echo=FALSE}
partialPlot(revenue.forest, green_test_noNA, green_rating)
```

We use our testing set to look at the partial effect of green rating. Green rating is a binary variable: we can see in the graph that being rated green is associated with an increase of roughly $80 in revenue per square foot per year. We can look to see if this partial effect is impacted by interaction with other relevant terms: Gas Costs, Electricity Costs and Size of the building.

```{r, echo=FALSE}
library(pdp)

partial(revenue.forest, pred.var = c("green_rating", "size"), plot=TRUE)

partial(revenue.forest, pred.var = c("green_rating", "Electricity_Costs"), plot=TRUE)

```

We can see that there is not a large partial effect of green rating among buildings that are the same size: this may indicate that our original $80 of increased revenue may be due to a confounder. Perhaps large buildings are more likely to be rated green and also more likely to have increased revenue per square foot per year. Additionally, there is little to no effect of being rated green if energy costs such as electricity are sufficiently low.
